Computing with cut systems +

### 2015-08-29 cutsystems-new.tex
New version of cut systems paper.

\documentclass[english,letter paper,12pt,leqno]{article}
\usepackage{stmaryrd}
\usepackage{amsmath, amscd, amssymb, mathrsfs, accents, amsfonts,amsthm}
\usepackage[all]{xy}
\usepackage{tikz}
\def\nicedashedcolourscheme{\shadedraw[top color=blue!22, bottom color=blue!22, draw=gray, dashed]}
\def\nicecolourscheme{\shadedraw[top color=blue!22, bottom color=blue!22, draw=white]}
\def\nicepalecolourscheme{\shadedraw[top color=blue!12, bottom color=blue!12, draw=white]}
\def\nicenocolourscheme{\shadedraw[top color=gray!2, bottom color=gray!25, draw=white]}
\def\nicereallynocolourscheme{\shadedraw[top color=white!2, bottom color=white!25, draw=white]}
\definecolor{Myblue}{rgb}{0,0,0.6}
\usepackage[a4paper,colorlinks,citecolor=Myblue,linkcolor=Myblue,urlcolor=Myblue,pdfpagemode=None]{hyperref}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

@@

	\theoremstyle{example}
	\newtheorem{definition}[theorem]{Definition}
	\newtheorem{example}[theorem]{Example}
	\newtheorem{remark}[theorem]{Remark}

@@

% Operators
\def\eval{\operatorname{ev}}
\def\res{\operatorname{Res}}
\def\Coker{\operatorname{Coker}}
\def\Ker{\operatorname{Ker}}
\def\im{\operatorname{Im}}
\def\can{\operatorname{can}}
\def\K{\mathbf{K}}
\def\D{\mathbf{D}}
\def\N{\mathbf{N}}
\def\LG{\mathcal{LG}}
\def\Ab{\operatorname{Ab}}
\def\Hom{\operatorname{Hom}}
\def\modd{\operatorname{mod}}
\def\Modd{\operatorname{Mod}}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\tot}{Tot}
\DeclareMathOperator{\ch}{ch}
\DeclareMathOperator{\str}{str}
\DeclareMathOperator{\hmf}{hmf}
\DeclareMathOperator{\HMF}{HMF}
\DeclareMathOperator{\hf}{HF}
\DeclareMathOperator{\At}{At}
\DeclareMathOperator{\Cat}{Cat}
\DeclareMathOperator{\Spec}{Spec}

@@

\title{Computing with cut systems}
\author{Daniel Murfet}

@@

@@

In this paper we develop a new semantics of composition using a bicategorical formalism that we call \emph{cut systems}, inspired by Gentzen's cut-elimination \cite{gentzen} and Girard's geometry of interaction \cite{girard_towards}. The formalism is used to ``make computable'' a bicategory of isolated hypersurface singularities and matrix factorisations that has been studied in the context of topological field theory \cite{McNameethesis, cr0909.4381}. This means that the operations making up the data of the bicategory are given algorithmically, and can be implemented on a computer. One of the main applications, to be given in a sequel \cite{cutsystems2}, is to construct a model of a fragment of Girard's linear logic \cite{girard_llogic}. Roughly speaking, this provides mathematical semantics for a programming language in which data types are interpreted as hypersurface singularities, and programs are interpreted as homotopy equivalence classes of matrix factorisations.

What is a semantics of composition? Consider that in a category or bicategory $\cat{A}$, for every triple of objects $a,b,c$ and morphisms
\[
\xymatrix@C+2pc{
a \ar[r]^-{\Phi} & b \ar[r]^-{\Psi} & c
}
\]
there is an assignment to $(\Psi, \Phi)$ of a composition $\Psi \circ \Phi$. A \emph{semantics of composition} for $\cat{A}$ is a mathematical construct providing some kind of inner structure for this ``assignment'' which is coherent as $a,b,c,\Phi,\Psi$ vary. The main examples of this kind of semantics are in computer science and mathematical logic.

For example, suppose that $\cat{A}$ is the category of sets and that we have functions
\[
\xymatrix@C+2pc{
\mathbb{N} \ar[r]^-{\Phi} & \mathbb{N} \ar[r]^-{\Psi} & \mathbb{N}
}
\]
where $\Phi$ and $\Psi$ are ``finitely described'' in the sense that they are computed by algorithms $X$ and $Y$, respectively. To be explicit, let us take $X,Y$ to be terms in the $\lambda$-calculus \cite{selinger}. If $[-]$ denotes assignment to an algorithm of the function it describes, then $\Phi = [X], \Psi = [Y]$ and $\Psi \circ \Phi = [Y \l X]$ where $Y \l X$ is the result of feeding the output of the algorithm $X$ as an input to the algorithm $Y$. In the logic literature this concatenation is known as a \emph{cut}.\footnote{We note that the notation $Y \l X$ for the cut is not standard, so don't blame the logicians for it.} To encounter the desired semantics we need the additional observation that the cut $Y \l X$ is not in normal form (whatever this means) but it may be reduced to normal form by a rewriting process on terms in the $\lambda$-calculus called $\beta$-reduction. This does not change the associated function: the normal form of $Y \l X$ also computes $\Psi \circ \Phi$.

A semantics of composition for the category of computable functions is then given by the process which begins with the pair $(Y,X)$ of finite descriptions of $\Psi,\Phi$, forms the cut $Y \l X$, and then computes a normal form giving a finite description of $\Psi \circ \Phi$.

This seems very far from geometry, but we can informally recast a problem of algebraic geometry in a similar light. Suppose $A,B,C$ are smooth projective algebraic varieties and that we are given triangulated functors between derived categories of coherent sheaves
\[
\xymatrix{
\mathbf{D}^b(A) \ar[r]^-{\Phi} & \mathbf{D}^b(B) \ar[r]^-{\Psi} & \mathbf{D}^b(C)
}\,.
\]
Suppose that $\Phi,\Psi$ are finitely described, in the sense that they are both integral functors associated to kernels: $\Phi$ is convolution with a complex of coherent sheaves $X$ on $A \times B$, and $\Psi$ is convolution with a complex $Y$ on $B \times C$. If for example $A,B,C$ are the $\operatorname{Proj}$ of graded algebras, $X$ and $Y$ can be described by lists of matrices of homogeneous polynomials. An obvious question is: what is the finite description of the functor $\Psi \circ \Phi$?

It is easy to see that $\Psi \circ \Phi$ is convolution with the kernel $Y \otimes^{\mathbb{L}}_{\cat{O}_B} X$, which is a complex of quasi-coherent sheaves on $A \times C$ with coherent cohomology, but this is not described by a finite amount of data. Finding a finite description of $\Psi \circ \Phi$ is precisely the problem of determing the complex of coherent sheaves which is isomorphic, in $\mathbf{D}^b(A \times C)$, to $Y \otimes^{\mathbb{L}}_{\cat{O}_B} X$. We refer to such a complex of coherent sheaves as a \emph{finite model} of $Y \otimes^{\mathbb{L}}_{\cat{O}_B} X$.

Consider the case where $A = C = \operatorname{Spec}(\mathbb{C})$, so $X,Y$ are complexes of coherent sheaves on $B$ and $Y \otimes^{\mathbb{L}}_{\cat{O}_B} X \cong \mathbb{R}\! \operatorname{Hom}_{\cat{O}_B}(Y^{\vee}, X)$ is a complex of infinite-dimensional vector spaces with finite-dimensional cohomology. An example of a finite model is an isomorphism in the derived category of all $\mathbb{C}$-vector spaces of this complex with its cohomology
\begin{equation}\label{eq:computing_hom_derived}
Y \otimes^{\mathbb{L}}_{\cat{O}_B} X \cong \bigoplus_{n \in \mathbb{Z}} \Ext^n(Y^{\vee},X)[-n]\,.
\end{equation}
One finite description of $Y \otimes^{\mathbb{L}}_{\cat{O}_B} X$, and transitively of $\Psi \circ \Phi$, is therefore a list of numbers: the dimensions of the vector spaces on the right hand side of \eqref{eq:computing_hom_derived}. Determining these numbers is difficult, to put it mildly: a fundamental problem in algebraic geometry is to understand how these numbers depend on the input data $X$ and $Y$.

By a (perhaps naive) analogy with computable functions, one might wonder if in the general case there exists an object $Y \l X$ in some triangulated category together with a reduction process that would help find a finite description for $\Psi \circ \Phi$. This is the question of whether or not there is a semantics of composition in the bicategory of smooth projective varieties and Fourier-Mukai kernels \cite{ct1007.2679}. Unfortunately, we do not know such a theory!

However, it turns out that there \emph{is} a satisfactory semantics of this kind when instead of smooth projective varieties we consider isolated hypersurface singularities $W,V,U$ and triangulated functors between homotopy categories of matrix factorisations \cite{yoshino98}
\begin{equation}\label{eq:pair_functors_hmf}
\xymatrix{
\hmf(W) \ar[r]^-{\Phi} & \hmf(V) \ar[r]^-{\Psi} & \hmf(U)
}\,.
\end{equation}
To explain further, we need to introduce the language of cut systems.

A \emph{cut system} $\cat{A}$ over a $\mathbb{Q}$-algebra $k$ has a class of objects and for each pair $a,b$ of objects a $k$-linear supercategory $\cat{A}(a,b)$. Informally, one should think of this as a category of finite models. There is an operation on $\cat{A}$, the \emph{cut}, which assigns to $X \in \cat{A}(a,b)$ and $Y \in \cat{A}(b,c)$ a representation $Y \l X$ in $\cat{A}(a,c)$ of a Clifford algebra which depends on the intermediate object $b$. The object $Y \l X$ and the endomorphisms giving the Clifford action are required to be effectively computable, say by explicit formulas, from the input data $X,Y$. In the example of a cut system developed in this paper, Clifford algebras arise as endomorphism algebras of exterior algebras, which in turn arise as the underlying graded modules of Koszul complexes. Associated to the cut system $\cat{A}$ is a bicategory $\cat{A}^\bullet$ in which the composition is derived from the cut operation, and so is also effectively computable.

A \emph{cut model} of a bicategory $\cat{B}$ is a cut system $\cat{A}$ with the same objects as $\cat{B}$, together with an equivalence $I: \cat{B} \lto \cat{A}^\bullet$. There are fully faithful functors $I: \cat{B}(a,b) \lto \cat{A}(a,b)$ which we think of as assigning finite models to the $1$-morphisms of $\cat{B}$, and an isomorphism of Clifford representations for every composable pair of $1$-morphisms $X,Y$ in $\cat{B}$, where $S$ denotes the spinor representation:
\begin{equation}\label{eq:intro_cut_semantics}
I(Y) \l I(X) \cong S \otimes_k I( Y \circ X )\,.
\end{equation}
The cut model provides $\cat{B}$ with a semantics of composition: given finite models $I(X), I(Y)$ of $X,Y$ there is a process beginning with the cut $I(Y) \l I(X)$ which extracts the subobject $I(Y \circ X)$ using the Clifford action. The result $I( Y \circ X )$ of this computation is the desired finite model of the composition $Y \circ X$ in $\cat{B}$.
\medskip

The formalism is adapted to, and motivated by, an example: the bicategory of Landau-Ginzburg models $\LG$. Let us return to the diagram \eqref{eq:pair_functors_hmf} as our starting point. There is an analogue of integral functors in this setting, so that it makes sense to talk about $\Phi$ and $\Psi$ as being respectively given by convolution with kernels $X,Y$. If $W \in k[x_1,\ldots,x_n]$ and $V \in k[z_1,\ldots,z_m]$ define isolated hypersurface singularities then $X$ is a finite rank matrix factorisation of $V - W$ over $k[x,z]$, and similarly for $Y$.

Again, the question is: what is the finite description of $\Psi \circ \Phi$? It is naturally given by convolution with a matrix factorisation $Y \otimes X$ which is not finitely generated: finding a finite description of $\Psi \circ \Phi$ therefore amounts to finding a finite model for $Y \otimes X$. To fit this into the formalism of cut systems, we use the  bicategory $\LG$ in which the objects are isolated hypersurface singularities\footnote{Roughly speaking. See Section \ref{section:superbicatLG}.} and $1$-morphisms $W \lto V$ are matrix factorisations of $V - W$. In this bicategory, the finite descriptions of the functors in \eqref{eq:pair_functors_hmf} give rise to a composable pair of $1$-morphisms
\begin{equation}\label{eq:cut_lg}
\xymatrix@C+2pc{
W \ar[r]^-{X} & V \ar[r]^-{Y} & U
}\,.
\end{equation}
whose composition in $\LG$ is the tensor product $Y \otimes X$.

The main results of this paper are the construction of a cut sytem $\cat{L}$ and cut model $I: \LG \lto \cat{L}^\bullet$. In the situation of \eqref{eq:cut_lg} $X$ and $Y$ are their own finite model (so $I(X) = X$ and $I(Y) = Y$) and a finite model of the composite $Y \otimes X$ in $\cat{B}$ may be computed from the cut $Y \l X$ in $\cat{L}$, which is an explicitly given finite rank matrix factorisation of $U - W$ together with a Clifford action described in terms of derivatives of the differentials of $X,Y$ and certain invariants called Atiyah classes.

One consequence is that $\LG$ is equivalent to the bicategory $\cat{L}^\bullet$, which is \emph{computable} in the sense that its operations may be embodied in software. To some extent this software already exists: a component of what we are here calling the cut system $\cat{L}$ (developed earlier in joint work with Dyckerhoff \cite{dm1102.2957}\footnote{For a more detailed discussion of the relation of this work with the present paper, see Remark \ref{remark:relation_to_toby_paper}.}) has already been implemented in the computer algebra package Singular in joint work with Carqueville \cite{khovhompaper}, and used to compute tables of Khovanov-Rozansky knot homologies. The code was also used in the discovery of the generalised orbifolding relations among ADE singularities by Carqueville, Ros Camacho and Runkel \cite{ade}.

\medskip
In conclusion, the purpose of this paper is to develop the theory of cut systems as a mathematical foundation for the application to be given in \cite{cutsystems2}. Let us briefly summarise the contents of that work: one can show that the superbicategory $\cat{L}^\bullet$ is monoidal, has adjoints and is pivotal. This makes it a natural target for models of Girard's linear logic, which is a generalisation of the $\lambda$-calculus in which the terms may be viewed as algorithms that have a refined syntax of resource usage compared to the $\lambda$-calculus. The model of linear logic to be given in \cite{cutsystems2} interprets algorithms $x,y$ as $1$-morphisms $[x],[y]$ in $\cat{L}^\bullet$.

This interpretation has two novelties: first of all, the target $\cat{L}^\bullet$ has a semantics of composition that may be compared to cut-elimination in linear logic (the analogue of $\beta$-reduction in the $\lambda$-calculus) as in Girard's geometry of interaction \cite{girard_towards}. This connection is what motivates the terminology of ``cut systems'' in the present paper. The second novelty is that, compared to the usual mathematical semantics of programming languages where algorithms are interpreted as functions between sets, our interpretation is bicategorical: in particular, equations $[x] \circ [y] \cong [y] \circ [x]$ or $[x]^2 \cong [x]$ on programs are represented by data that is part of the semantics.

%. For a precise definition see Section \ref{??}}. To such a $W$ we associate a category of algebraic objects called matrix factorisations. A matrix factorisation of $W(x)$ is a matrix of polynomials $D$ satisfying $D^2 = W \cdot I_n$, where $I_n$ is the identity matrix.

%For $W = x y^2 + y^3$ an example is
%\[
%D_1 = \begin{pmatrix} 0 & x - i y \\ x + iy & 0 \end{pmatrix} %, \qquad D_2 = \begin{pmatrix} 0 & y - x \\ (y - e^{2 \pi i /3} x)(y - e^{4 \pi i /3} x ) & 0 \end{pmatrix}
%\]
%A \emph{morphism} of matrix factorisations $\Phi: D_1 \lto D_2$ is a matrix of polynomials $\Phi$ satisfying the relation $\Phi D_1 = D_2 \Phi$. Together the matrix factorisations of $W$ and the morphisms between them form the category of matrix factorisations of $W$.

%In the second part of the paper we explain how the multiplicative fragment of linear logic can be interpreted in the bicategory of Landau-Ginzburg models, with formulas mapped to isolated hypersurface singularities and proofs mapped to matrix factorisations. The multiplicative fragment is not very expressive, so our focus in the present paper is not on embedding sophisticated programs into $\cat{LG}$. Rather, we explain how cut systems give a dynamic model of cut-elimination following Girard's geometry of interaction. The basic construction of a cut system involves homological perturbation, and by interpreting this as a fixed point construction we will show that the result of normalization can be expressed in a power series expansion similar to Girard's famous execution formula.

@@

By $\hmf(k[x],W)^{\oplus}$ we denote the full subcategory of $\HMF(k[x],W)$ consisting of matrix factorisations which are direct summands (in the homotopy category) of finite rank matrix factorisations. Since $\HMF(k[x],W)$ is idempotent complete, there is an equivalence (see Remark \ref{remark:idempotent_completion} for the notation)
\[
\hmf(k[x],W)^\omega \lto \hmf(k[x],W)^{\oplus}
\]
sending a pair $(X,e)$ to an infinite rank matrix factorisation splitting the idempotent $e$.

@@

The bicategory of Landau-Ginzburg models $\LG$ has for its objects pairs $(x,W)$ consisting of an ordered set of variables $x = (x_1,\ldots,x_n)$ and a potential $W \in k[x] = k[x_1,\ldots,x_n]$. Given potentials $W \in k[x]$ and $V \in k[z]$ the supercategory $\cat{LG}(W,V)$ is
\[
\cat{LG}(W,V) = \hmf(k[x,z], V - W)^{\oplus}\,.
\]
Given a third potential $U \in k[y]$ the composition law is a functor
\[
\LG(W,V) \otimes_k \LG(V, U) \lto \LG(W, U)
\]
which sends a pair of $1$-morphisms $X: W \lto V$ and $Y: V \lto U$ to the tensor product
\begin{equation}\label{eq:tensor_comp}
Y \otimes X = ( Y \otimes_{k[z]} X, d_{Y \otimes X} = d_Y \otimes 1 + 1 \otimes d_X )\,.
\end{equation}
This statement requires some care: $k[x,y,z]$ is an infinite rank free module over $k[x,z]$ and so $Y \otimes X$ is an infinite rank matrix factorisation of $U - W$ over $k[x,z]$. However one can prove that $Y \otimes X$ is a direct summand of a finite rank matrix factorisation \cite{dm1102.2957} and therefore a valid object in $\LG(W,U)$.

@@

@@

In a cut system the composition rule produces objects which are representations of Clifford algebras. We briefly recall the basic theory of these algebras and their modules; for a full discussion see \cite{friedrich}. For $n \ge 0$ the Clifford algebra $A_n$ is the associative $\mathbb{Z}_2$-graded $k$-algebra generated by odd elements $a_1,\ldots,a_n$ and $a_1^\dagger, \ldots, a_n^\dagger$ subject to the following relations
\[
[a_i, a_j] = 0, \qquad [a_i^\dagger, a_j^\dagger] = 0, \qquad [a_i, a_j^\dagger] = \delta_{ij}\,.
\]
for all $1 \le i, j \le n$. In this paper all our commutators are graded, $[a,b] = ab - (-1)^{|a||b|} ba$. We set $A_0 = k$. The algebras $A_n$ are Morita trivial in the sense that they are isomorphic to the endomorphism algebras of finite rank free $\mathbb{Z}_2$-graded $k$-modules.

Indeed, if $F_n = k \theta_1 \oplus \cdots \oplus k \theta_n$ with $|\theta_i| = 1$ for all $i$, then

\begin{lemma} With $S_n = \bigwedge F_n$ the map $A_n \lto \End_k(S_n)$ defined by
\[
a_i \mapsto \theta_i \wedge (-), \qquad a_i^\dagger \mapsto \theta_i^*\, \lrcorner\, (-)
\]
is an isomorphism of $\mathbb{Z}_2$-graded $k$-algebras, where $\lrcorner$ denotes contraction.
\end{lemma}

In particular this means that every $\mathbb{Z}_2$-graded $A_n$-module is isomorphic to $S_n \otimes_k V$ for some $\mathbb{Z}_2$-graded $k$-module $V$. From the isomorphism of $k$-modules
\[
S_m \otimes_k S_n = \bigwedge F_m \otimes_k \bigwedge F_n \cong \bigwedge F_{m+n} = S_{m+n}
\]
we deduce an isomorphism of algebras
\begin{equation}\label{eq:algebra_A_additive}
A_m \otimes_k A_n \cong A_{m+n}\,.
\end{equation}
Given $m, n \ge 0$ we introduce notation for the $A_m$-$A_n$-bimodule
\begin{equation}
S_{m,n} = S_m \otimes_k S_n^*
\end{equation}
where $S_n^* = \Hom_k(S_n, k)$. There is an isomorphism of $A_l$-$A_n$-bimodules
\begin{equation}\label{eq:isosbimodule}
S_{l,m} \otimes_{A_m} S_{m,n} = S_l \otimes_k S_m^* \otimes_{A_m} S_m \otimes_k S_n^* \cong S_{l,n}\,.
\end{equation}
In the case $k = \mathbb{C}$, $S_n$ is often referred to as the \emph{spinor representation} \cite[p.14]{friedrich}.

>>>>

% TODO
% - fix the equivalence of L with LG
% - add stuff to the example

\documentclass[english,letter paper,12pt,leqno]{article}
\usepackage{stmaryrd}
\usepackage{amsmath, amscd, amssymb, mathrsfs, accents, amsfonts,amsthm}
\usepackage[all]{xy}
\usepackage{dsfont}
\usepackage{tikz}
\def\nicedashedcolourscheme{\shadedraw[top color=blue!22, bottom color=blue!22, draw=gray, dashed]}
\def\nicecolourscheme{\shadedraw[top color=blue!22, bottom color=blue!22, draw=white]}
\def\nicepalecolourscheme{\shadedraw[top color=blue!12, bottom color=blue!12, draw=white]}
\def\nicenocolourscheme{\shadedraw[top color=gray!2, bottom color=gray!25, draw=white]}
\def\nicereallynocolourscheme{\shadedraw[top color=white!2, bottom color=white!25, draw=white]}
\definecolor{Myblue}{rgb}{0,0,0.6}
\usepackage[a4paper,colorlinks,citecolor=Myblue,linkcolor=Myblue,urlcolor=Myblue,pdfpagemode=None]{hyperref}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{setup}[theorem]{Setup}

@@

	\theoremstyle{example}
	\newtheorem{definition}[theorem]{Definition}
	\newtheorem{example}[theorem]{Example}
	\newtheorem{remark}[theorem]{Remark}
	\newtheorem{strat}[theorem]{Strategy}

@@

% Operators
\def\eval{\operatorname{ev}}
\def\res{\operatorname{Res}}
\def\Coker{\operatorname{Coker}}
\def\Ker{\operatorname{Ker}}
\def\im{\operatorname{Im}}
\def\can{\operatorname{can}}
\def\K{\mathbf{K}}
\def\D{\mathbf{D}}
\def\N{\mathbf{N}}
\def\LG{\mathcal{LG}}
\def\Ab{\operatorname{Ab}}
\def\Hom{\operatorname{Hom}}
\def\modd{\operatorname{mod}}
\def\Modd{\operatorname{Mod}}
\def\be{\begin{equation}}
\def\ee{\end{equation}}
\def\nN{\mathds{N}}
\def\nZ{\mathds{Z}}
\def\nQ{\mathds{Q}}
\def\nR{\mathds{R}}
\def\nC{\mathds{C}}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\tot}{Tot}
\DeclareMathOperator{\ch}{ch}
\DeclareMathOperator{\str}{str}
\DeclareMathOperator{\hmf}{hmf}
\DeclareMathOperator{\HMF}{HMF}
\DeclareMathOperator{\hf}{HF}
\DeclareMathOperator{\At}{At}
\DeclareMathOperator{\Cat}{Cat}
\DeclareMathOperator{\Spec}{Spec}

@@

\title{Computing with cut systems +}
\author{Daniel Murfet}

@@

\tableofcontents

@@

The bicategory of Landau-Ginzburg models $\LG$ has polynomials $W(x)$ called \textsl{potentials} as its objects, and $1$-morphisms $W(x) \lto V(y)$ are matrix factorisations of the difference $V(y) - W(x)$ over $\nC[x,y]$. This bicategory has been studied extensively in connection with topological field theory \cite{??} and knot invariants \cite{??}. The most important and nontrivial structure in this bicategory is the functor
\be\label{eq:intro_functor_comp}
\LG(V,U) \otimes_{\nC} \LG(W, V) \lto \LG(W,U)
\ee
which composes $1$-morphisms for potentials $W(x),V(y),U(z)$. Continuing the approach developed with Dyckerhoff in \cite{??}, the aim of this paper is to explicitly describe this functor. Any increase in our understanding of the operation of composition in $\LG$ has implications for the subjects where this bicategory makes an appearance, but our primary application is to the semantics of linear logic constructed in \cite{??}.

\begin{itemize}
\item We define another bicategory without units $\cat{L}$
\item We prove $\cat{L}$ is equivalent to $\LG$
\item We show how to model $\cat{L}$ in schemes (next paper)
\end{itemize}

First of all, let us be more precise about the kind of explicit description we are looking for. Observe that once we have chosen a homogeneous basis, a matrix factorisation of $W(x) - V(y)$ is just a matrix of polynomials
\[
D \in M_{2r}( \nC[x,y] ) \qquad \text{ satisfying } \qquad D^2 = (W(x) - V(y)) \cdot I_{2r}\,.
\]
If we fix the rank $r$ and bound the degrees of the monomials appearing in $D$ by some integer $k$, it is easy to parametrise such matrix factorisations by the closed points of an affine scheme $\mathscr{M}_{r,k}(W,V)$. Then the set of objects of $\LG(W,V)$ can be described as a disjoint union of inductive limits of these schemes
\[
\operatorname{ob} \LG(W,V) = \bigsqcup_{r \ge 1} \bigcup_{k \ge 0} \mathscr{M}_{r,k}(W,V)\,.
\]
The kind of explicit description we are looking for is a lifting of the function \eqref{eq:intro_functor_comp} on the level of objects to a morphism of schemes
\be\label{eq:composition_as_schememap}
\mathscr{M}_{r',k'}(V,U) \times \mathscr{M}_{r,k}(W,V) \lto \mathscr{M}_{f(r,r'),g(k,k')}(W, U)\,.
\ee
Unfortunately this appears difficult, if not impossible, as stated. In this article we explain how the replacement of $\LG(W,U)$ by an equivalent category $\cat{L}(W,U)$, called the \textsl{Clifford thickening}, allows for such a lifting of \eqref{eq:intro_functor_comp} to a morphism of schemes.

% One consequence is that $\LG$ is equivalent to the bicategory $\cat{L}^\bullet$, which is \emph{computable} in the sense that its operations may be embodied in software. To some extent this software already exists: a component of what we are here calling the cut system $\cat{L}$ (developed earlier in joint work with Dyckerhoff \cite{dm1102.2957}\footnote{For a more detailed discussion of the relation of this work with the present paper, see Remark \ref{remark:relation_to_toby_paper}.}) has already been implemented in the computer algebra package Singular in joint work with Carqueville \cite{khovhompaper}, and used to compute tables of Khovanov-Rozansky knot homologies. The code was also used in the discovery of the generalised orbifolding relations among ADE singularities by Carqueville, Ros Camacho and Runkel \cite{ade}. 
\medskip

@@

By $\hmf(k[x],W)^{\oplus}$ we denote the full subcategory of $\HMF(k[x],W)$ consisting of matrix factorisations which are direct summands (in the homotopy category) of finite rank matrix factorisations. Since $\HMF(k[x],W)$ is idempotent complete, there is an equivalence (see Remark \ref{remark:idempotent_completion} for the notation)
\be\label{eq:omegavsoplus}
\hmf(k[x],W)^\omega \lto \hmf(k[x],W)^{\oplus}
\ee
sending a pair $(X,e)$ to an infinite rank matrix factorisation splitting the idempotent $e$.

@@

The bicategory of Landau-Ginzburg models $\LG$ has for its objects pairs $(x,W)$ consisting of an ordered set of variables $x = (x_1,\ldots,x_n)$ and a potential $W \in k[x] = k[x_1,\ldots,x_n]$. Given potentials $W \in k[x]$ and $V \in k[y]$ the supercategory $\cat{LG}(W,V)$ is
\[
\cat{LG}(W,V) = \hmf(k[x,y], V - W)^{\oplus}\,.
\]
Given a third potential $U \in k[z]$ the composition law is a functor
\[
\LG(W,V) \otimes_k \LG(V, U) \lto \LG(W, U)
\]
which sends a pair of $1$-morphisms $X: W \lto V$ and $Y: V \lto U$ to the tensor product
\begin{equation}\label{eq:tensor_comp}
Y \otimes X = ( Y \otimes_{k[y]} X, d_{Y \otimes X} = d_Y \otimes 1 + 1 \otimes d_X )\,.
\end{equation}
This statement requires some care: $k[x,y,z]$ is an infinite rank free module over $k[x,z]$ and so $Y \otimes X$ is an infinite rank matrix factorisation of $U - W$ over $k[x,z]$. However one can prove that $Y \otimes X$ is a direct summand of a finite rank matrix factorisation \cite{dm1102.2957} and therefore a valid object in $\LG(W,U)$.

@@

\subsection{Partial derivatives as homotopies}\label{section:partial}

Let $R = k[x_1,\ldots,x_n]$ be a polynomial ring and $X$ a finite rank matrix factorisation of $W$ over $R$. Let us fix a homogeneous basis for $X$ so that we may identify $d_X$ with a matrix, and differentiate this matrix entry-wise to obtain the matrix $\lambda_i = \partial_{x_i}(d_X)$.

\begin{lemma} The odd $R$-linear operator $\lambda_i$ on $X$ is independent, up to homotopy, of the choice of basis.
\end{lemma}

\begin{lemma} As operators on $X$ we have $[ d_X, \lambda_i ] = \partial_{x_i} W \cdot 1_X$.
\end{lemma}
\begin{proof}
Applying the Leibniz rule to the equation $d_X d_X = W$ we obtain
\be\label{eq:partials_are_homotopies}
d_X \partial_{x_i}(d_X) + \partial_{x_i}(d_X) d_X = \partial_{x_i} W
\ee
as claimed.
\end{proof}

It follows that any element in the Jacobi ideal $( \partial_{x_1}W , \ldots, \partial_{x_n} W )$ acts null-homotopically on any morphism in the homotopy category $\hmf(R,W)$. Moreover these homotopies $\lambda_i$ are central, in the following sense:

\begin{lemma}\label{lemma:naturalityoflambda} Let $\phi: X \lto X'$ be a morphism of matrix factorisations and write $\lambda_i = \partial_{x_i}(d_X)$ and $\lambda'_i = \partial_{x_i}(d_{X'})$. Then there is a homotopy $\phi \lambda_i \simeq \lambda'_i \phi$.
\end{lemma}
\begin{proof}
From $d_{X'} \phi = \phi d_X$ we deduce
\[
\partial_{x_i}( d_{X'} ) \phi + d_{X'} \partial_{x_i}( \phi ) = \partial_{x_i}( \phi ) d_X + \phi \partial_{x_i}( d_X )
\]
and hence
\[
\partial_{x_i}( d_{X'} ) \phi - \phi \partial_{x_i}( d_X ) = \partial_{x_i}(\phi) d_X - d_{X'} \partial_{x_i}(\phi)
\]
as claimed.
\end{proof}

\begin{lemma}\label{lemma:commutator_of_lambdas} There is a homotopy $[ \lambda_i, \lambda_j ] \simeq \partial_{x_i x_j}(W) \cdot 1_X$.
\end{lemma}
\begin{proof}
Differentiating \eqref{eq:partials_are_homotopies} with respect to $j$ leaves
\[
\partial_{x_j}(d_X) \partial_{x_i}(d_X) + d_X \partial_{x_jx_i}(d_X) + \partial_{x_jx_i}(d_X) d_X + \partial_{x_i}(d_X) \partial_{x_j}(d_X) = \partial_{x_ix_j}W
\]
from which we read off
\[
\lambda_i \lambda_j + \lambda_j \lambda_i = \partial_{x_ix_j}W + [d, -\partial_{x_jx_i}(d_X)]\,.
\]
\end{proof}

@@

In a cut system the composition rule produces objects which are representations of Clifford algebras. We briefly recall the basic theory of these algebras and their modules; for a full discussion see \cite{friedrich}. For $n \ge 0$ the Clifford algebra $C_n$ is the associative $\mathbb{Z}_2$-graded $k$-algebra generated by odd elements $\gamma_1,\ldots,\gamma_n$ and $\gamma_1^\dagger, \ldots, \gamma_n^\dagger$ subject to the following relations
\be\label{eq:clifford_relations}
[\gamma_i, \gamma_j] = 0, \qquad [\gamma_i^\dagger, \gamma_j^\dagger] = 0, \qquad [\gamma_i, \gamma_j^\dagger] = \delta_{ij}\,.
\ee
for all $1 \le i, j \le n$. In this paper all our commutators are graded, $[a,b] = ab - (-1)^{|a||b|} ba$. We set $C_0 = k$. These Clifford algebras have essentially only one nontrivial representation, sometimes called the \emph{spinor representation} \cite[p.14]{friedrich}. To define this representation we use the $\nZ_2$-graded $k$-module
\[
F_n = k \theta_1 \oplus \cdots \oplus k \theta_n
\]
where the $\theta_i$ are formal variables of odd degree, and set
\be
S_n = \bigwedge F_n = \bigwedge\big( k \theta_1 \oplus \cdots \oplus k \theta_n \big)\,.
\ee

\begin{definition}\label{defn:contraction} Left multiplication in the exterior algebra defines an odd operator
\begin{gather*}
\theta_i \wedge (-): S_n \lto S_n\,,\\
\theta_{j_1} \cdots \theta_{j_r} \mapsto \theta_i \theta_{j_1} \cdots \theta_{j_r}\,.
\end{gather*}
Contraction from the left defines an odd operator
\begin{gather*}
\theta_i^*\, \lrcorner\, (-): S_n \lto S_n\,,\\
\theta_{j_1} \cdots \theta_{j_r} \mapsto \sum_{l=1}^r (-1)^{(l-1)} \delta_{i,j_l} \theta_{j_1} \cdots \widehat{ \theta_{j_l} } \cdots \theta_{j_r}\,.
\end{gather*}
\end{definition}

Usually we write simply $\theta_i$ for the operator $\theta_i \wedge (-)$ and $\theta_i^*$ for the operator $\theta_i^*\, \lrcorner\, (-)$. The spinor representation is defined by the next lemma, which also shows that the algebras $C_n$ are Morita trivial in the sense that they are isomorphic to the endomorphism algebras of a finite rank free $\mathbb{Z}_2$-graded $k$-module.

\begin{lemma}\label{defn:spinorrep} The map $C_n \lto \End_k(S_n)$ defined by
\[
\gamma_i \mapsto \theta_i \wedge (-), \qquad \gamma_i^\dagger \mapsto \theta_i^*\, \lrcorner\, (-)
\]
is an isomorphism of $\mathbb{Z}_2$-graded $k$-algebras.
\end{lemma}
